\documentclass{article} %<<<
\usepackage{amsmath}
\usepackage{fancyvrb}
\usepackage{mathpazo}
\usepackage{microtype}
\usepackage{tikz}
\usepackage{xcolor}

\usepackage{hyperref}

\usetikzlibrary{automata,positioning,arrows}
\tikzset{automaton/.style={auto,node distance=1cm,on grid}}
\tikzset{every state/.style={minimum size=0pt,inner sep=1pt}}
\tikzset{transition/.style={->,>=stealth',shorten >=1pt}}
\tikzset{every initial by arrow/.style={transition}}
\tikzset{initial text={}}

\title{Static Checking of TOPL Properties}
\author{
  Dino Distefano
  \and Radu Grigore
  \and Rasmus Lerchedahl Petersen
  \and Nikos Tzevelekos}

\newcommand{\infer}[2]{\frac{\displaystyle\;#1\;}{\displaystyle\;#2\;}}
\newcommand{\3}[3]{\{\,#1\,\}\;#2\;\{\,#3\,\}}

\newcommand{\noterg}[2]{\textcolor{gray}{[\textcolor{red}{#1}: #2]}}
\newcommand{\rg}[1]{\noterg{rg}{#1}}

%>>>
\begin{document} %<<<
\maketitle

We want to prove that a given program does not violate a given topl property.
The main idea is that almost any static analyzer is sufficient.
Why?
Well, suppose the \emph{runtime} monitor faults when a violation occurs.
If a static analyzer concludes that the instrumented program, including the monitor, cannot fault, then this means the topl property was checked statically.
The static analyzer must be able to modularly infer specifications, for example via abduction.
Otherwise the verification would be very inconvenient.

Consider the topl property
\[\begin{tikzpicture}[automaton,node distance=2.7cm]\footnotesize
  \node[state,initial] (q0) [label=below:$0$] {};
  \node[state] (q1) [label=below:$1$,right=of q0] {};
  \node[state] (q2) [label=below:$2$,right=of q1] {};
  \node[state] (q3) [label=below:$3$,right=of q2] {};
  \node[state,accepting] (q4) [label=below:$4$,right=of q3] {};
  \path[transition]
    (q0) edge node{{\it I}{\sf :=}{\it C}{\sf.}{\tt iterator}{\sf()}} (q1)
         edge[loop above] node{\tt*} ()
    (q1) edge node{{\it J}{\sf :=}{\it c}{\sf.}{\tt iterator}{\sf()}} (q2)
         edge[loop above] node{\tt*} ()
    (q2) edge node{{\it i}{\sf.}{\tt remove}{\sf()}} (q3)
    (q3) edge node{{\it j}{\sf .}{\tt next}{\sf()}} (q4);
\end{tikzpicture}\]
To check that a program does not violate it we run a static verifier on a modified program.
We add global variables {\tt state}, {\tt c}, {\tt i}, and~{\tt j}.
We replace calls to {\tt iterator} by calls to {\tt iterator'}, calls to {\tt remove} by calls to {\tt remove'}, and calls to {\tt next} by calls to {\tt next'}.
The primed methods are implemented as follows.
\begin{Verbatim}[fontsize=\footnotesize]
  Iterator iterator'() {
    emit(new Event(1, new Object[]{this}));
    Iterator r = iterator();
    emit(new Event(2, new Object[]{r}));
    return r;
  }
  void remove'() {
    emit(new Event(3, new Object[]{this}));
    remove();
  }
  Object next'() {
    emit(new Event(4, new Object[]{this}));
    return next();
  }
\end{Verbatim}
This is essentially the same as is done in the runtime verifier---each observable method announces when it is called and when it returns.
Each call and each return gets an \emph{event identifier} (from~$1$ to~$4$ in this example).
In this example events carry $0$~or~$1$ objects, but for other properties there might be more, which is why an array is used to store them.
The method {\tt emit} uses another global variable, the queue~$E$ of events.
\begin{Verbatim}[fontsize=\footnotesize]
  void emit(Event e) {
    E.enque(e);
    step();
    assert state != 4;
  }
\end{Verbatim}
The method {\tt step} advances the configuration of the r-topl automaton, which is represented by the integer {\tt state} and the queue {\tt events}.

We should find a specification for {\tt step}.
Alternatively, the static analyzer could infer the specification from an implementation of {\tt step}.
But, of course, we cannot expect a generic specification inference heuristic to do as well as an algorithm that exploits the special structure of~{\tt step}.

The specification of {\tt step} is complex, so good notation is needed.
The queue~$E$ has bounded size~$E_{\rm sz}$.
In this example, $E_{\rm sz}$~is $0$,~$1$, or~$2$.
Each event $E_i$ contains a bounded number of values~$E_{i,j}$, and has an identifier~$E_{i,{\rm id}}$.
Thus, the queue is represented by a set of primitive global variables.
In this example the variables are $E_{\rm sz}$,~$E_{0,{\rm id}}$, $E_{0,0}$, $E_{1,{\rm id}}$, and~$E_{1,0}$.

Let $g$ be a guard and $a$ be an action.
Guards are formulas interpreted over a pair~$(\sigma,\ell)$ of a store~$\sigma$ and a letter~$\ell$.
The store $\sigma$ consists of a few global variables ($c$,~$i$, and~$j$ in this example).
The letter $\ell$ is the oldest unprocessed event~$E_0$.
Actions are sequences of simple assignments
\[ x_1:=y_1;\;\ldots;\;x_n:=y_n\]
and are seen as substitutions.
Within postconditions of Hoare triples we may use the special unary function~${\sf old}$.
Occurrences within occurrences of ${\sf old}(\ldots)$ are said to be \emph{old occurrences}; the other are \emph{new occurrences}.
For example, in $x={\sf old}(x)$, the variable~$x$ has one new occurrence and one old occurrence.
Actions are seen as substitutions.
\begin{align*}
&(x:=y)\circ g &&\text{is obtained from $g$ by substituting $y$ for the new occurences of $x$} \\
&(x:=y)\bullet g &&\text{is obtained from $g$ by substituting $y$ for the old occurrences of $x$}
\end{align*}
Substitutions are applied starting with the last assignment.
\begin{align*}
(a_1;a_2)\circ g &= a_1\circ(a_2\circ g) \\
(a_1;a_2)\bullet g &= a_1\bullet(a_2\bullet g)
\end{align*}
When parentheses are omitted, they are assumed to be as above.

Each state~$q$ has an associated formula which means `state $q$ is active and we know enough events to simulate one step of the automaton'
\[ P_q \quad=\quad ({\it state}=q) \land (E_{\rm sz}\ge d_q) \]
where $d_q$~is the maximum length of a transition outgoing from~$q$.
Each transition~$t$ has two associated formulas.
Its \emph{precondition} identifies the configurations in which $t$~is enabled.
\begin{multline*}
P_t \quad=\quad
    g_0
    \land\bigl(a_0;(E_0:=E_1)\circ g_1\bigr)
    \land\ldots\\\ldots
    \land\bigl(a_0;(E_0:=E_1);a_1;(E_0:=E_2);\ldots;a_{n-2};(E_0:=E_{n-1})\circ g_{n-1}\bigr)
\end{multline*}
Here $g_0,\ldots,g_{n-1}$ are the guards of~$t$, and $a_0,\ldots,a_{n-1}$ are its actions.
Its \emph{postcondition} encodes the action of the transition.
\[ Q_t \quad=\quad
  ({\it state}=q')
  \land D(n)
  \land \bigl(a_0;(E_0:=E_1);a_1;\ldots;(E_0:=E_{n-1});a_{n-1}\bullet I\bigr)
  \]
The state $q'$ is the destination of~$t$.
The formula~$D(n)$ encodes dropping $n$~events from~$E$.
The formula~$I$ states that the variables $x_1$,~$x_2$, $\ldots$~that make up the store remain unchanged.
\begin{align*}
D(n) \quad&=\quad
  \bigl(E_{\rm sz}={\sf old}(E_{\rm sz})-n\bigr)
  \land\bigl(E_0={\sf old}(E_n)\bigr)
  \land\bigl(E_1={\sf old}(E_{n+1})\bigr)
  \land\ldots\\
I \quad&=\quad
  \bigl(x_1={\sf old}(x_1)\bigr)
  \land\bigl(x_2={\sf old}(x_2)\bigr)
  \land\ldots
\end{align*}
In particular, $D(0)$~is the identity for the variables that represent the queue~$E$.
In our example the following definitions would work:
{\def\=#1{(#1={\sf old}(#1))}
\begin{align*}
D(0) \quad&=\quad \={E_{\rm sz}}\land\={E_0}\land\={E_1} \\
D(1) \quad&=\quad (E_{\rm sz}={\sf old}(E_{\rm sz})-1)\land(E_0={\sf old}(E_1)) \\
D(2) \quad&=\quad (E_{\rm sz}={\sf old}(E_{\rm sz})-2) \\
I \quad&=\quad \=c\land\=i\land\=j
\end{align*}}

Let $t_1$~and~$t_2$ be transitions outgoing from~$q$.
Let $S$ be the symbolic state just before the call to {\tt step}.
Suppose $S$ implies both $P_{t_1}$~and~$P_{t_2}$.
Intuitively this means that $t_1$~and~$t_2$ are both enabled.
After the call to {\tt step} the concrete state should satisfy $Q_{t_1}\lor Q_{t_2}$.
Nondeterminism is represented by disjunction.
(Conjunction cannot work because $({\it state}=0)\land({\it state}=1)$ is inconsistent.)
In general, let $t_1$,~$t_2$, \dots,~$t_n$ be all the transitions outgoing from state~$q$.
Then the following are $2^{n-1}$ good contracts for {\tt step}:
\[
  \left\{\begin{aligned}
  &P_q&&\\
  &P_{t_k}  &&\text{for $k\in T$} \\
  &\lnot P_{t_k}  &&\text{for $k\notin T$}
  \end{aligned}\right\}
  \quad\text{\tt step}\quad
  \left\{\bigvee_{k\in T}Q_{t_k}\right\}
  \qquad\text{for $\emptyset\subset T\subseteq \{1,2,\ldots,n\}$}
\]
The conditions listed in the precondition are implicitly conjoined.
The set~$T$ of enabled transitions must be not empty.
Otherwise, if no outgoing transition is enabled, the contract must drop one event.
\[
  \{P_q\land\lnot P_{t_1}\land\ldots\land\lnot P_{t_n}\}
  \quad\text{\tt step}\quad
  \{({\it state}=q)\land D(1)\land I\}
\]
If the queue~$E$ does not contain sufficient events, then {\tt step} should be a no-operation.
\[
  \{({\it state}=q)\land(E_{\rm sz}<d_q)\}
  \quad\text{\tt step}\quad
  \{({\it state}=q)\land D(0)\land I\}
\]

Note that the preconditions listed so far for {\tt step} are mutually unsatisfiable.
This means that the conjunction rule of Hoare logic cannot be used.
The disjunction rule of Hoare logic, on the other hand, must be used, to handle the disjunctions that represent nondeterminism.

Let us now spell out the contract of {\tt step} in the case of the running example.
There are four no-op contracts with satisfiable preconditions:
\begin{align*}
\{({\it state}=0)\land(E_{\rm sz}<2)\} \quad\text{\tt step}\quad
  \{({\it state}=0)\land D(0)\land I\} \\
\{({\it state}=1)\land(E_{\rm sz}<2)\} \quad\text{\tt step}\quad
  \{({\it state}=1)\land D(0)\land I\} \\
\{({\it state}=2)\land(E_{\rm sz}<1)\} \quad\text{\tt step}\quad
  \{({\it state}=2)\land D(0)\land I\} \\
\{({\it state}=3)\land(E_{\rm sz}<1)\} \quad\text{\tt step}\quad
  \{({\it state}=3)\land D(0)\land I\}
\end{align*}
The most complicated state is probably $1$:
It has two outgoing transitions of different lengths, and the longer one uses both (nontrivial) guards and actions.
The loop on~$1$ has the precondition ${\it true}$ and the postcondition
\[ Q_{1\to1} \quad=\quad ({\it state}=1) \land D(1) \land I \]
The transition from~$1$ to~$2$ has the precondition
\[ (E_{0,{\rm id}}=1)\land(c=E_{0,0}) \]
which is essentially the guard of the first step, and the postcondition
\[ Q_{1\to2} \quad=\quad
({\it state}=2)\land D(2)\land (c={\sf old}(c))\land (i={\sf old}(i))
  \land (j={\sf old}(E_{1,0}))\]
For two transitions, there are $2^2-1$ contracts that encode rollback configuration transitions.
\begin{align*}
&\bigl\{({\it state}=1)\land(E_{\rm sz}\ge2)\land
  \lnot\bigl((E_{0,{\rm id}=1})\land(c=E_{0,0})\bigr)\bigr\}
  \quad\text{\tt step}\quad
  \bigl\{Q_{1\to1}\bigr\} \\
&\bigl\{({\it state}=1)\land(E_{\rm sz}\ge2)\land
  (E_{0,{\rm id}=1})\land(c=E_{0,0})\bigr\}
  \quad\text{\tt step}\quad
  \bigl\{Q_{1\to1}\lor Q_{1\to2}\bigr\} \\
&\bigl\{{\it false}\bigr\}\quad\text{\tt step}\quad\bigl\{Q_{1\to2}\bigr\}
\end{align*}
The contract that encodes the skip rollback transitions is
\[ \bigl\{{\it false}\bigr\}\quad\text{\tt step}\quad
  \bigl\{({\it state}=1)\land D(0)\land I\bigr\} \]
The ${\it false}$ preconditions in the contract appear whenever the precondition of the loop $1\to1$ is negated.

In summary, each state with $n$ outgoing transitions contributes three types of triples for {\tt step}:
\begin{itemize}
\item $2^n-1$ triples that simulate rollback configuration transitions
\item $1$ triple that simulates skip configuration transitions
\item $1$ no-op triple
\end{itemize}

\medskip

So far we have a method to synthesize the specification of {\tt step}.
What we need is a way to change a given specification for {\tt iterator}, {\tt next}, and in general all methods that are mentioned in the topl property.
As an intermediate step towards that goal we should synthesize the specification of {\tt emit}.

Recall the rule of composition of Hoare logic.
\[\infer
  {\3{P}{C}{Q}\quad\3{Q}{D}{R}}
  {\3{P}{C;D}{R}}
\]
The postconditions we synthesized earlier use a special function {\sf old}, which may not appear in preconditions.
The rule of composition, however, demands that the same formula~$Q$ is both a postcondition and a precondition.
It seems difficult to synthesize a sound and simple rule of composition that interacts well with the special function {\sf old}.
It is easier to remove {\sf old} from triples before composing them.
By definition, ${\sf old}(f(t_1,\ldots,t_n))$ is the same as $f({\sf old}(t_1),\ldots,{\sf old}(t_n))$, and
\[
  \3{P}{C}{Q} \quad\text{if and only if}\quad
  \3{P*(x'=x)}{C}{({\sf old}(x):=x')\circ Q}
\]
when $x'$~does not occur in~$P,Q$.
In general we handle a program $C_1;\ldots;C_n$ whose steps modify a variable~$x$ by introducing versions $x^{(1)}$, \dots,~$x^{(n)}$.

Now we go back to the method {\tt emit}.
The implementation of {\it enque\/} is inlined.
\begin{align*}
\{\,\,\}
  \quad&E[E_{\rm sz}]:=e
  \quad\{\,E_{\rm sz}\stackrel E\mapsto e\,\} \\
\{\,\,\}
  \quad&E_{\rm sz}:=E_{\rm sz}+1
  \quad\{\,E_{\rm sz}:={\sf old}(E_{\rm sz})+1\,\} \\
\{\,P_{\rm step}\,\}\quad&E,\sigma,{\it state}:={\it step}()\quad
  \{\,Q_{\rm step}\,\} \\
\{\,{\it state}\ne4\,\}\quad&{\sf assert}\;{\it state}\ne4\quad\{\,\,\}
\end{align*}
This time the variables that are modified are listed explicitly.
The first two statements modify a few of the variables of the queue~$E$.
Remember that $E$~stands for a group of variables.
The other variables are unchanged.
Recall now the frame rule.
\[\infer
  {\3{P}{C}{Q}}
  {\3{P*F}{C}{Q*F}}
  \quad\text{if ${\it free}(F)$ and ${\it assignable}(C)$ are disjoint}
\]
As warmup, we compose the two assignments that implement {\it enque}.
First, we rewrite the second triple in a form without {\sf old}, by introducing a fresh logical variable~$E^{(2)}_{\rm sz}$.
(The subscript is~$2$ because it is the second statement of {\it emit}.)
\[\infer
  {\3{}{E_{\rm sz}:=E_{\rm sz}+1}{E_{\rm sz}={\sf old}(E_{\rm sz})+1}}
  {\3
    {E^{(2)}_{\rm sz}=E_{\rm sz}}
    {E_{\rm sz}:=E_{\rm sz}+1}
    {E_{\rm sz}=E^{(2)}_{\rm sz}+1}}
  \]
Second, the frame rule gives us
\[\infer
  {\3{}{E[E_{\rm sz}]:=e}{E_{\rm sz}\stackrel E\mapsto e}}
  {\3
    {E^{(2)}_{\rm sz}=E_{\rm sz}}
    {E[E_{\rm sz}]:=e}
    {(E_{\rm sz}\stackrel E\mapsto e)*(E^{(2)}_{\rm sz}=E_{\rm sz})}} \]
and
\[\infer
  {\3
    {E^{(2)}_{\rm sz}=E_{\rm sz}}
    {E_{\rm sz}:=E_{\rm sz}+1}
    {E_{\rm sz}=E^{(2)}_{\rm sz}+1}}
  {\3
    {(E^{(2)}_{\rm sz}=E_{\rm sz})*(E^{(2)}_{\rm sz}\stackrel E\mapsto e)}
    {E_{\rm sz}:=E_{\rm sz}+1}
    {(E_{\rm sz}=E^{(2)}_{\rm sz}+1)*(E^{(2)}_{\rm sz}\stackrel E\mapsto e)}}
\]
The postcondition of the first statement is now equivalent (but not identical) to the precondition of the the second statement.
The rule of composition gives us a triple for {\it enque}.
\[\3
  {E^{(2)}_{\rm sz}=E_{\rm sz}}
  {E[E_{\rm sz}]:=e; E_{\rm sz}:=E_{\rm sz}+1}
  {(E_{\rm sz}=E^{(2)}_{\rm sz}+1)*(E^{(2)}_{\rm sz}\stackrel E\mapsto e)} \]

\medskip

The next step would be to find a triple for the last two statements of {\it emit}, and then to find a triple for {\it emit}.
Before that, let us clarify whether $E$~is an array or simply a group of variables.
The two are, of course, inter-changable, so the question is only which is more convenient for our purposes.
For example, the assignment $E[E_{\rm sz}]:=e$ is easily handled if $E$ is regarded to be an array, and the frame rule is used to infer that the other cells remain unchanged.
If $E$~was a group of variables, then that assignment would have to be considered as syntactic sugar for
\begin{align*}
      &{\bf if}\;E_{\rm sz}=0\;{\bf then}\;E_0:=e \\
  ;\; &{\bf if}\;E_{\rm sz}=1\;{\bf then}\;E_1:=e \\
      &\vdots
\end{align*}
which would lead to considerably uglier formulas.
\rg{Continue here.}

\medskip


Let us now compose the last two statements of {\it emit}, a {\it step\/} followed by the assertion that the error state was not reached.
First, we need to get rid of {\sf old}.

\rg{Update:}
We cannot frame the triple of {\it step\/} with ${\it state}\ne4$, because {\it state\/} is assignable by {\it step}.
But, we can frame the triple of the assertion with $Q_{\rm step}$.
The precondition $Q_{\rm step}*({\it state}\ne4)$ is in some cases equivalent to $Q_{\rm step}$, in some cases equivalent to {\it false\/}, and no other case is possible, because of the special form of~$Q_{\rm step}$, as synthesized above.
Thus, we simply avoid synthesizing the postconditions of {\it step\/} that contain ${\it state}=4$.
For all other cases the rule of composition gives the following triple.
\[\3
  {P_{\rm step}}
  {E,\sigma,{\it state}:={\it step}(); {\sf assert}\;{\it state}\ne4}
  {Q_{\rm step}} \]

\end{document} %>>>
% vim:fmr=<<<,>>>:
